---
title: "Simulating fluctuations in estimated CFR with sample size"
author: "Thibaut Jombart"
date: "`r Sys.Date()`"
output:
  rmdformats::material:
    highlight: kate
    css: !expr here::here('css', 'style.css')
link-citations: yes
---


```{r settings, echo = FALSE}
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 5,
  collapse = TRUE
)
```



# Proof of concept

Simulating the outcomes of cases using a known case fatality ratio (CFR) using
**R** is easy. This process is by definition a Binomial draw, which is
implemented by the function `rbinom`. See `?rbinom` for details of how this
function works. The arguments of the function are simple:

```{r dbinom}
args(rbinom)
```

* `n` is the number of times we want to draw from the Binomial distribution
* `prob` is the probability, here, of death
* `size` is the size of the sample, here, the number of cases

So that if we want to simulate the number of death amongst 12 cases, with a CFR
of 40%, we would then use something like:

```{r simple_simul}

# set the random seed
set.seed(1)

# do the 'experiment' once
rbinom(n = 1, size = 12, prob = 0.4)

# repeat that 'sampling' 20 times
rbinom(n = 20, size = 12, prob = 0.4)

# do it 1 million times, and examine the resulting distribution
x <- rbinom(n = 1e6, size = 12, prob = 0.4)
barplot(table(x),
        col = "#7A9548",
        border = "white",
        xlab = "Number of deaths",
        ylab = "Frequency")

```



# Making our own function

The following function wraps up the principle above, with the added tweaks:

* arguments have a clearer epidemiological meaning
* the output reports the number of cases, the number of deaths, survivors, and
  the associated ratio
  
```{r simulate_outcome}

# Define the new function

simulate_outcome <- function(n = 1,
                             n_cases,
                             cfr) {
  ## simulate outcomes using binomial sampling
  res <- rbinom(n = n,
                prob = cfr,
                size = n_cases)
  
  ## shape final data.frame
  ### note: as the last object is returned, we do not need to assign the created
  ### data.frame to any object
  data.frame(cases = n_cases,
             deaths = res,
             survivors = n_cases - res,
             ratio = res / n_cases)
}


# try it
simulate_outcome(1, 100, 0.2)
simulate_outcome(10, 60, 0.33)

```




# Simulating estimated CFR over time

To show how sample size affects the estimation of CFR, we contrast the following
situations, all assuming a true CFR of 20% over a period of 90 days:

* 5 daily cases
* 10 daily cases
* 100 daily cases

*dplyr* and *ggplot2* will be used to reorganise and display the results


```{r simulations}
library(tidyverse)

# run simulations
sim_5_cases <- simulate_outcome(90, n_cases = 5, cfr = 0.2)
sim_10_cases <- simulate_outcome(90, n_cases = 10, cfr = 0.2)
sim_100_cases <- simulate_outcome(90, n_cases = 100, cfr = 0.2)

# bind them into a single data.frame
sims <- bind_rows(sim_5_cases, sim_10_cases, sim_100_cases) %>%
  mutate(cases = factor(cases)) %>%
  group_by(cases) %>% 
  mutate(day = 1:n()) %>% # add 'day' variable
  select(day, everything()) # reorder data to have day first
glimpse(sims)

# plot results
ggplot(sims, aes(x = day, y = ratio)) +
  theme_bw() +
  geom_line(color = "#DF7416") +
  facet_grid(cases ~ .) +
  geom_hline(yintercept = 0.2, color = "grey", linetype = 2) +
  labs(title = "Simulating fluctuations in of estimated CFR",
       x = "Day", y = "Estimated CFR",
       subtitle = "for 5, 10, and 100 daily cases - true CFR: 0.2")

```
